NMRlipids IV: Headgroup \& glycerol backbone structures, and cation binding in bilayers with PS lipids

### 2018-12-11 Manuscript/manuscriptPS.tex
Update manuscript

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids interact with signaling and other proteins via
electrostatic interactions and direct binding, and induce
membrane fusion and phase separation together with calcium ions.
Molecular details of these phenomena are not well understood
because accurate models to interpret the experimental data has not
been available. Here, we collect a set of experimental NMR data which
can be used together with molecular dynamics (MD) simulations
to interpret the lipid headgroup structures and details of ion binding
in pure PS and mixed PS:PC lipid bilayers. We use the open collaboration
method to collect data from available MD simulation models of PS lipids.
None of the models reproduce the NMR data with experimental accuracy,
but the best models suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
all the tested force fields overestimate the cation binding affinity to lipid bilayers containing
PS lipids. In contrast to PC lipids, the response of PS headgroups
to the bound ions qualitatively differs from experiments in the tested MD
simulation models. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe negatively charged membranes and their interactions with ions.
The work is performed in the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

@@

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18}.
However, the headgroup structures and ion binding affinity 
predicted by the MD simulation force fields for bilayers containing PS lipids are not
evaluated against experiments. Recent studies comparing the
C-H bond order parameters between simulations and experiments show that the
structure of PC lipid headgroup and glycerol backbone are not well
captured by most simulation models \cite{botan15} and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}. Based on this data,
the cation binding affinity to POPC bilayer was then improved by implicitly
including the electronic polarizability using the electronic continuum correction \cite{melcr18}.

\begin{table*}[htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{The list of MD simulations of pure PS bilayers without additional salt.
  Simulation details are given in the supplementary information
    %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{PSsystems}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c r r r r r c c}
  %\hline
  % some footnotes are not visible in typeset-MS (pdf)
 lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
    \hline
    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
    POPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
  \hline
    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
    \end{tabular}
%\end{minipage}
%\end{sidewaystable*} 
\end{table*}

@@

\subsection{Solid state NMR experiments}

@@

{\emph{R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{S-DROSS experiment}. The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{NMR numerical simulations}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

@@

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{nmrlipids.blogspot.fi}) and
the GitHub repository (\url{github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in 
Tables \ref{PSsystems} (pure PS systems without additional ions) 
and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
Simulation details are given in the SI.
The simulation data is also indexed in the
searchable database (\url{nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}

@@

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available. 
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}
\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup and glycerol backbone from experiments
    and simulations with different force fields.
    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data is without additional ions.
    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
    Light blue areas spanning 0.04 units around the average of the extremal experimental values,
    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
    The vertical bars shown for some simulation values
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

@@

The headgroup and glycerol backbone of PS lipids show wide variety
in the order parameters and structures (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS})
between different simulation models,
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids in the previous study
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

However, the best performing models,
Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} b) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

The observed conformations of PS headgroup may be useful when interpreting experiments and
can guide the further force field development. However, more accurate simulations are
required to confirm the results because the simulated PS headgroup order parameters
are not within the experimental error bars in any of the tested models.

@@

\begin{figure}[!htb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\cite{ferreira16}.
  }
  \todo{Simulation of CHARMM36 at 298K should be maybe rerun with Gromacs 5.} \\
  \todo{The data from POPC used in Gromos-CKP by would be useful for this plot.}
\end{figure}

>>>>

\begin{abstract}
  % insert abstract here
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids interact with signaling and other proteins via
electrostatic interactions and direct binding, and induce
membrane fusion and phase separation together with calcium ions.
Molecular details of these phenomena are not well understood
because accurate models to interpret the experimental data has not
been available. Here, we collect a set of experimental NMR data which
can be used together with molecular dynamics (MD) simulations
to interpret the lipid headgroup structures and details of ion binding
in pure PS and mixed PS:PC lipid bilayers. We use the open collaboration
method to collect data from available MD simulation models of PS lipids.
None of the models reproduce the NMR data with experimental accuracy,
but the best models suggest that the carboxyl group in the serine headgroup
does not rotate freely. In line with the previous results for PC lipids,
none of the tested force fields correctly captures the cation binding affinity to lipid bilayers containing
PS lipids. In contrast to PC lipids, the response of PS headgroups
to the bound ions qualitatively differs from experiments in the tested MD
simulation models. The collected experimental dataset and simulation
results pave the way for improvement of lipid force fields to correctly
describe negatively charged membranes and their interactions with ions.
The work is performed in the NMRlipids open collaboration project (\url{nmrlipids.blogspot.fi}).
\end{abstract}

@@

%\label{}
\section{Introduction}
Phosphatidylserine (PS) is the most common negatively
charged lipid in eykaryotic membranes.
PS lipids compose 8.5\% of total lipid weight of erythrocytes,
but the abundance varies between different organelles up to
25-35\% in plasma membrane \cite{lemmon08,leventis10,li14}.
Despite of the relatively low abundance, PS lipids
are important signaling molecules. They interact with
signaling proteins \cite{leventis10}, regulate
surface charge and protein localization \cite{yeung08}, and
induce protein aggregation \cite{zhao04,gorbenko06}.
Some protein domains spesifically interact PS lipids,
while others are attracted by general electrostatics and the
binding can be regulated by calcium \cite{leventis10}.
Therefore, the structural details
of lipid headgroups and the details of cation binding
are crucial for the PS mediated signaling processes.

@@

\begin{table*}[htb]
%\begin{sidewaystable*}[!p]
\centering
\caption{The list of MD simulations of pure PS bilayers without additional salt.
  Simulation details are given in the supplementary information
    %   The lipid force fields named as in our previous work~\cite{botan15}.
}\label{PSsystems}
%begin{minipage}[t]{\textwidth}
\begin{tabular}{l c r r r r r c c}
  %\hline
  % some footnotes are not visible in typeset-MS (pdf)
 lipid/counter-ions & force field for lipids / ions & \footnote{Number of lipid molecules with largest mole fraction}N$_{\rm l}$   &  \footnote{Number of water molecules}N$_{\rm w}$  & \footnote{Number of additional cations}N$_{\rm c}$  & \footnote{Simulation temperature}T (K)  & \footnote{Total simulation time}t$_{{\rm sim}}$(ns) & \footnote{Time used for analysis}t$_{{\rm anal}}$ (ns) &   \footnote{Reference for simulation files}files\\
    \hline
    POPS/Na$^+$  & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298K} \\
    POPS/K$^+$   & CHARMM36 \cite{venable13} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36POPS298Kpotassium} \\
    POPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{charmm36uaPOPS298K} \\
    POPS/Na$^+$  & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 500 & 100  & \cite{macrogPOPS298Kcorrect} \\
    POPS/K$^+$   & MacRog \cite{maciejewski14}  & 128 & 4480 & 0  & 298  & 200 & 150 & \cite{macrogPOPS298KwithK} \\
    POPS/Na$^+$  & lipid17  \cite{gould18} / JC  \cite{joung08}   & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSjcions} \\
    POPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 298  & 600 & 100 & \cite{lipid17POPSff99ions} \\
    POPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}          & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{bergerPOPS298K} \\
    POPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp1POPS303K} \\
    POPS/Na$^+$  & GROMOS-CKP \cite{??} \todoi{Correct citation(s) for CKP.}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{ckp2POPS303K} \\
    POPS/Na$^+$  & Slipids \cite{jambeck13}  & 128 & 4480 & 0  & 298  & 500 & 100 & \cite{slipidsPOPS298K} \\
  \hline
    DOPS/Na$^+$  & CHARMM36 \cite{venable13}       & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36DOPS303K} \\
    DOPS/Na$^+$  & CHARMM36ua \cite{??} \todoi{Correct citation for CHARMMua DOPS}   & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{charmm36uaDOPS303K} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / JC  \cite{joung08}    & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSjcions} \\
    DOPS/Na$^+$  & lipid17 \cite{gould18} / ff99 \cite{aqvist90}  & 128    & 4480   & 0   & 303  & 600 & 100 & \cite{lipid17DOPSff99ions} \\
    DOPS/Na$^+$  & Berger \cite{mukhopadhyay04,??}    & 128  & 4480  & 0  & 303  & 500 & 100 & \cite{bergerDOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKPM \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp1DOPS303K} \\
    DOPS/Na$^+$  & GROMOS-CKP  \cite{??} \todoi{Correct citation(s) for CKP.} & 128 & 4480 & 0  & 303  & 500 & 100 & \cite{ckp2DOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 128 	& 4480  & 0  & 303  & 500 & 100 & \cite{slipidsDOPS303K} \\
    DOPS/Na$^+$  & Slipids \cite{jambeck13}        & 288 	& 11232 & 0  & 303  & 200 & 100 & \cite{slipidsDOPSfiles} \\
    \end{tabular}
%\end{minipage}
%\end{sidewaystable*} 
\end{table*}

Several classical molecular dynamics (MD) simulation studies are done
to understand PS headgroups, their influence on lipid bilayer properties and
interactions with
ions \cite{cascales96,pandit02,mukhopadhyay04,pedersen06,vernier09,boettcher11,molina12,jurkiewicz12,venable13,pan14,vangaveti14,melcrova16,valentine18,hallock18}.
However, the results strongly depend on the used force field parameters.
For example, the recent simulations using NBfix parameters for calcium \cite{kim16} in
CHARMM36 force field \cite{klauda10,venable13} combined with 2D infrared spectroscopy
suggests that calcium ions interacts only with the carboxylate group of PS lipids \cite{valentine18}, while
the same force field without the NBfix parameters together with the NMR chemical shifts and
REDOR experiments suggests a significant binding affinity also to the phosphate region \cite{hallock18}.
On the other hand, simulations with the Berger force field \cite{berger97,mukhopadhyay04}
combined with fluorescent and vibrational sum frequency spectroscopy suggested a significant
calcium binding also to the carbonyls in the acyl chains \cite{melcrova16}. 
We have recently demonstrated that such controversies can be resolved by
comparing C-H bond order parameters of lipid headgroups between simulations
and experiments \cite{botan15,catte16}, which can be directly
measured from NMR experiments with high accuracy and compared to simulations
in order to evaluate the simulation model quality or to interpret the experiments \cite{ollila16}.
Previous studies showed that the structure of PC lipid headgroup and glycerol backbone are not well
captured by most simulation models \cite{botan15} and that the cation binding to PC
lipid bilayers is overestimated \cite{catte16}. Based on this data, the cation binding
affinity to POPC bilayer was then improved by implicitly including the electronic polarizability
using the electronic continuum correction \cite{melcr18}.

@@

\subsection{C-H bond order parameters from the natural abundance $^{13}$C NMR}

@@

{\emph{C-H bond order parameters from the R-PDLF experiment}}. The parameters are described according to Figures 1c and 2c of the original reference
for the R-PDLF experiment~\cite{dvinskikh04}.  The refocused-INEPT delays $\tau_1$ and $\tau_2$ were 1.94 ms and 0.97 ms, respectively.
Radio frequency pulses with the nutation frequencies: 46.35 kHz (R18$^7_1$ pulses), 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$),
50 kHz (SPINAL64 $^1$H decoupling pulses). The $t_1$ increment was equal to 10.79 $\mu$s $\times18\times2$ and 32 points in the indirect
dimension were recorded using 1024 scans for each, with recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{Order parameter signs from the S-DROSS experiment}.
The parameters are described according to Figures 1b and 1c of the original reference for the S-DROSS
experiment~\cite{gross97}. The refocused-INEPT delay $\delta_2$ was 1.19 ms. The $\tau_1$ and $\tau_2$ in the S-DROSS recoupling
blocks $R$ were set as 39.4 $\mu$s and 89.4 $\mu$s, respectively. Radio frequency pulses with the nutation
frequencies: 63.45 kHz ($^{13}$C 90$^{\rm{o}}$ and 180$^{\rm{o}}$), 50 kHz ($^1$H SPINAL64 decoupling).
The $t_1$ increment (dipolar recoupling dimension) was 800 $\mu$s and a total of 8 points along $t_1$ were
measured using 1024 scans for each with a recycle delay of 5 s and a spectral with of 149.5 ppm.

\emph{Numerical simulations of S-DROSS curves.}
The numerical simulations of S-DROSS curves were performed with the SIMPSON simulation package \cite{bak00} using the $^{13}$C--$^1$H
dipolar couplings determined by the R-PDLF experiments or calculated from the known $^2$H quadrupolar couplings \cite{browning80} as input.
The chemical shift anisotropy and homonuclear couplings were neglected, and the input file {\it{rep2000}} was used to simulate the random
distribution of bilayer orientations in the samples studied.

@@

\subsection{Molecular dynamics simulations}
Molecular dynamics simulation data was collected using
the Open Collaboration method \cite{botan15}.
The NMRlipids project blog (\url{www.nmrlipids.blogspot.fi}) and
the GitHub repository (\url{www.github.com/NMRLipids/NMRlipidsIVotherHGs})
were used as the communication platforms.
The simulated systems are listed in 
Tables \ref{PSsystems} (pure PS systems without additional ions) 
and \ref{mixedIONsystems} (mixed PC:PS systems with various ions concentrations).
Further simulation details are given in the SI.
The simulation data is also indexed in the
searchable database (\url{www.nmrlipids.fi}),
and in the NMRlipids/MATCH GitHub repository (\url{https://github.com/NMRLipids/MATCH}).

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.5cm]{../Figs/fig1_POPS.pdf}
  \caption{\label{PShgSIGNSsimpson}
    The headgroup and glycerol backbone region of the (A) INEPT spectrum and
    (B) 2D R-PDLF spectra.
    (C) Experimental SDROSS data (points) and SIMPSON simulations (blue lines) with
    the order parameter values of -0.12 for the $\beta$-carbon, and 0.09 and -0.02
    for the $\alpha$-carbon slittings. The S-DROSS curve from SIMPSON simulation with positive value
    for the smaller $\alpha$-carbon order parameter (dashed grey).
  }
  \todo{I think that the peak labeling would be good to show also in (A).}
\end{figure}
\begin{figure}[!htb]
  \centering
  \includegraphics[width=9.0cm]{../Figs/PCPScomp.eps}
  \caption{\label{HGorderParameters}
    (A) Chemical structures and labels for the headgroup and glycerol backbone carbons.
    (B) Headgroup and glycerol backbone order parameters of POPS measured in this work compared
    with the previously published values from DOPS ($^2$H NMR, 0.1M of NaCl) \cite{browning80} and 
    POPC  ($^{13}$C NMR) \cite{ferreira13} experiments. Signs of the PS order parameters
    are measured in this work. Signs of the PC order parameters are measured previously~\cite{ferreira16}.
  }
\end{figure}

@@

\begin{figure}[]
  \centering
  \includegraphics[width=9.0cm]{../Figs/HGorderparametersPS.pdf}
  \caption{\label{HGorderParametersPS}
    Order parameters of PS headgroup and glycerol backbone from experiments
    and simulations with different force fields.
    Experimental data for DOPS is measured with 0.1~M of NaCl~\cite{browning80},
    while all the other data is without additional ions.
    All the data for DOPS is at 303~K and all the data for POPS is at 298~K.
    Light blue areas span 0.04 units around the average of the extremal experimental values,
    which is the expected quantitative accuracy of experimental values~\cite{ollila16}.
    The vertical bars shown for some simulation values
    are not error bars, but demonstrate that for these systems
    we had at least two data sets; the ends of the bars mark the extreme values
    from the sets, and the dot marks their measurement-time-weighted average.
  }
\end{figure}

The headgroup and glycerol backbone order parameters of 
POPS measured in this work are in good agreement with the previously reported
values from $^2$H NMR experiments of DOPS \cite{browning80} (Fig. \ref{HGorderParameters}).
When compared with the previously measured values for POPC \cite{ferreira13} (Fig. \ref{HGorderParameters}),
the $\beta$-carbon order parameter is significantly more negative and $\alpha$-carbon
experiences a significant forking (different order parameters for two hydrogens in the same carbon \cite{ollila16}) in the PS headgroup.
These features have been intepreted to arise from a rigid PS headgroup
conformation, stabilized by hydrogen bonds or electrostatic
interactions \cite{browning80,buldt81}, but detailed structrural interpretation is not
available.

\subsection{Headgroup and glycerol backbone in simulations of PS lipid bilayers without additional ions}

@@

The headgroup and glycerol backbone of PS lipids show wide variety in the order parameters and structures
between different simulation models (Figs. \ref{HGorderParametersPS} and \ref{HGandGLYstructuresPS}),
as previously observed also for PC lipids \cite{botan15}.
The models perform generally less well for PS lipids than for PC lipids 
(Figs.~\ref{HGorderParametersPS} and \ref{comparisonTablePS} vs. Figs. 2 and 4 in Ref.~\cite{botan15}).
Therefore, the interpretation of structural differences between PC and PS headgroups
from simulations is not straightforward.

However, the best performing models,
Slipids, CHARMM36 and CHARMM36ua, reproduce the larger forking of the $\alpha$-carbon 
and the Slipids model reproduces also the lower of the $\beta$-carbon order parameter
when comparing the PS results to PC (Fig. \ref{HGorderParametersPS} vs. Fig. 2 in Ref.~\citenum{botan15}).
Interestingly, the C$_\alpha$-C$_\beta$-C$_\gamma$-O$_\gamma$  dihedral with
a single and narrow peak in the angle distribution close to 120$^{\circ}$ is more restricted
in the best three models than in other models which give two maxima with different angles (Fig. \ref{dihedralsHG}).
The restricted motion is also
visible in the sampled conformations (Figs.~\ref{HGstructuresPSandPC} (A) and \ref{HGandGLYstructuresPS})
suggesting that the rotation of carboxyl group is limited in the serine headgroup.
In addition, the N-C$_\beta$-C$_\alpha$-O$_\alpha$ dihedral exhibit a more asymmetric
and restricted angle distribution for PS than for PC headgroup in
CHARMM36 simulations in best agreement with experiments
(Figs.~\ref{HGstructuresPSandPC} (B) and \ref{dihedralsHGpc}).
The results might manifest the increased rigidity anticipated in the early experimental studies \cite{browning80,buldt81}.
Also, the sampled conformations of glycerol backbone
significantly vary between different simulation models 
(Figs. \ref{dihedralsGLY}, \ref{HGandGLYstructuresPS}, \ref{dihedralsGLYpc} and \ref{HGandGLYstructuresPSPC}),
but further analysis is beyond the scope of this work which the PS headgroup.

The suggested characteristic conformations of PS headgroup may be useful when interpreting experiments and
can guide the further force field development. However, more accurate simulations are
required to confirm the suggested conformations because the simulated PS headgroup order parameters
are not within the experimental error bars in any of the tested models.

@@

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPCvsPS.eps}
  \caption{\label{HGorderparametersPCvsPS}
    Changes of POPC headgroup order parameters with increasing amount of POPS in POPC:POPS mixtures at 298~K.
    Experimental values are from Ref. \citenum{scherer87} with the signs measured in Ref.~\citenum{ferreira16}.
  }
  \todo{After we know which force field is used for POPC in Gromos-CKP simulations, we might be able to
    add Gromos-CKP data into this plot.}
\end{figure}

\begin{figure}[!tb]
  \centering
  \includegraphics[width=8.0cm]{../Figs/HGorderparametersPSvsPC.eps}
  \caption{\label{HGorderparametersPSvsPC}
    Changes of POPS headgroup order parameters with increasing amount of POPC in POPC:POPS mixtures at 298~K.
    Experimental values with the signs are measured for pure POPS system in this work.
    The signs are assumed to be the same for the mixture and the values are from Ref. \citenum{roux90}.
%    Since the experimental data of POPS in pure and diluted mixture come from
%    different experimental sets ($^{13}$C NMR in this work and $^{2}$H NMR in Ref. \citenum{roux90}),
%    the experimental change of the order parameter is less accurate than in typical measurements
%    where same technique is used in all conditions, see discussion about qualitative and quantitative 
%    accuracy in Ref. \citenum{ollila16}.
    The y-axis for the $\alpha$-carbon results of POPS (bottom) is transferred
    with the same value for both order parameters such that the lower order
    parameter value from pure POPS is at zero to correctly illustrate the significant forking.
  }
\end{figure}
